---
title: 'Reproducible Research: Peer Assessment 1'
author: "Bence Szikora"
date: "2025-02-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
I used the dplyr and lattice packages for this assignment

```{r}
#load library
library(dplyr)
library(lattice)

#Load data
data<-read.csv( unzip('activity.zip'),header = TRUE)
```

## What is mean total number of steps taken per day?

```{r}
#ignore NA values
clean_data <- data %>%
        filter(!is.na(steps))

#Calculate the total number of steps per day
steps_per_day <- clean_data %>%
        group_by(date) %>%
        summarise(total_steps=sum(steps))

#mean and median of the total number of steps taken per day

mean_steps<-mean(steps_per_day$total_steps)
print(paste("Mean total steps per day:",mean_steps))
median_steps<-median(steps_per_day$total_steps)
print(paste("Median total steps per day:",median_steps))

```

Histogram of the total number of steps taken each day:

```{r}
#make a histogram
hist(steps_per_day$total_steps, col="grey", 
     main="Histogram of total steps per day",
     xlab="Total steps per day",
     ylab="Frequency", border="black", breaks=20,xlim = c(0,25000))
```

## What is the average daily activity pattern?

A time series plot of the average number of steps taken:

```{r}
#calculate the average number of steps taken, averaged across all days

mean_steps_per_int <- clean_data %>%
        group_by(interval) %>%
        summarise(mean_steps=mean(steps))

#make a time series plot

plot(type="l",x=mean_steps_per_int$interval,y=mean_steps_per_int$mean_steps,
     main="Average daily activity pattern",
     xlab="Daily intervals", 
     ylab="Average steps per intervals across all days",
     col="black")
```

The 5-minute interval that, on average, contains the maximum number of steps:

```{r}
#Get the interval with the maximum number of steps on average
max_value<-mean_steps_per_int$interval[which.max(mean_steps_per_int$mean_steps)]
print(max_value)
```

## Imputing missing values

Code to describe and show a strategy for imputing missing data

```{r}
#Calculate the total number of missing values
mv<-sum(is.na(data$steps))
print(paste("The total number of missing values in the dataset:",mv))

#Filling the NA values with the mean of that 5 minute interval

data_imp <- data %>%
        left_join(mean_steps_per_int, by="interval") %>%
        mutate(filled_steps= ifelse(is.na(steps),mean_steps,steps))%>%
        select(-mean_steps)

#make a new dataframe with the filled data and the same structure

new_data <- data_imp[,c("filled_steps",'date','interval')]
```

Histogram of the total number of steps taken each day after missing values are imputed

```{r}
#Make a histogram of the total number of steps taken each day
steps_per_day_new <- new_data %>%
        group_by(date) %>%
        summarise(total_steps=sum(filled_steps))

#histogram
hist(steps_per_day_new$total_steps, col="lightblue", 
     main="Histogram of total steps per day with filled data",
     xlab="Total steps per day",
     ylab="Frequency", border="black", breaks=20,xlim = c(0,25000))
```

```{r}
#Calculate and report the mean and median total number of steps taken per day.
mean_steps_2<-mean(steps_per_day_new$total_steps)
print(paste("Mean total steps per day (flled data):",mean_steps_2))
median_steps_2<-median(steps_per_day_new$total_steps)
print(paste("Median total steps per day (filled data):",median_steps_2))
```

## Are there differences in activity patterns between weekdays and weekends?

Panel plot comparing the average number of steps taken per 5-minute 
interval across weekdays and weekends

```{r}
#Generate the weekday/weekend factor
new_data <- new_data %>%
        mutate(days_type= ifelse(weekdays(as.POSIXlt(new_data$date)) %in% c("szombat","vasárnap"),"weekend","weekday"))
new_data$days_type=as.factor(new_data$days_type)

#calculate the average of steps on weekday and weekend per interval
new_data_int <- new_data %>%
        group_by(days_type,interval)%>%
        summarise(mean_steps=mean(filled_steps),.groups="drop")

#plot with lattice
xyplot(mean_steps~interval|days_type, data=new_data_int, layout=c(1,2), type="l", 
       main="A time series plot of the 5-minute interval per average of steps on weekend and weekday",
       xlab = "Interval",ylab="Number of steps")
```

